A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-22 tex/codatadp.tex
more edits, including new Tab. 1 data

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcal{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcal{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\AST(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\kern.45em \parallel \kern.45em}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
%\newcommand{\interp}[2]{#1\llbracket #2 \rrbracket}
%%% @ANDY: I prefer a notation that puts more emphasis on #2 (which is, in essence,
%%% the object that is returned) and less on #1
%\newcommand{\interp}[2]{\llbracket #2 \rrbracket_{#1}}
%%% And one that doesn't look so semantic
%\newcommand{\interp}[2]{\lceil #2 \rceil_{#1}}
\newcommand{\interp}[2]{\left< \smash{#2} \right>_{#1}}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
%\newcommand{\nf}[1]{{{#1}{\downarrow}}}
\newcommand{\nf}[1]{\lfloor#1\rfloor}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

The $\mu$-term $u$
is a \emph{self-similar subterm} of~$t$ if
$u$ is a strict subterm of $t$,
$t$ is of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$,
$u$ it is of the form $\MU y.\; \const{C}( u_1, \ldots, u_n )$,
and $\interp{t}{ t_k } \aequiv \interp{t}{ u_k }$ for all $k \in [n]$.
The term $t$ is \emph{normal} if it does not contain a self-similar subterm
and all of its strict subterms are also normal.
For example, the $\mu$-term $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal,
since $\MU y.\; \const{C}( y )$ is a self-similar subterm of $t$.
In particular, note that their arguments have the same expansion with respect to $t$,
that is,
$\interp{t}{\MU y.\; \const{C}( y )} =$
$\MU y.\; \const{C}( \interp{t}{y}^{\{y\}} ) =$
$\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$\interp{t}{y} =$
$\MU y.\; \const{C}( \interp{t}{y}^{\{y\}} ) =$
$\MU y.\; \const{C}( y )$.
The $\mu$-term $t$ of the form $\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a self-similar subterm of $t$,
noting that
$\interp{t}{\MU y.\; \const{C}( x )} =
\MU y.\; \const{C}( \interp{t}{x}^{\{y\}} ) =
\MU y.\; \const{C}( \interp{t}{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\{y\}} ) =
\MU y.\; \const{C}( \MU x.\; \const{C}( \interp{t}{\MU y.\; \const{C}( x )}^{\{x,y\}} ) ) =
\MU y.\allowbreak\; \const{C}( \MU x.\; \const{C}( y ) )$
is $\alpha$-equivalent to $\interp{t}{x} = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.
For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$, we can
recursively construct its \emph{normal form} $\nf{t}$,
by replacing all of the self-similar subterms of $t$ with $x$
%and by replacing all of the abnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For example, $\nf{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )} = \MU x.\; \const{C}( x )$.

@@

In the following, we consider only normal models.
When constructing a model $\M$ for $\Ec$,
it remains only to define how $\M$ interprets wrongly applied selector terms and variables.
For the latter, this will be based on the mapping $\Val$ constructed in phase~2 of the calculus.
We compute the \emph{completion} $\Val^\AST$ of $\Val$ for normal model $\M$
by assigning values from $\M$ to unassigned variables in the domain of $\Val$.

@@

We construct $\Val^\AST$ by exhaustively applying the following rule to $\nf{\Val}$:
\[
\hbox{\(
\inferrule{
  {\Varec{x}}\typ{\tau} \in \FV( \Val )
  \quad
  \MU {\Varec{x}}.\; t \aequiv v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_{\Varec{x}}(\Val)
}{
  \Val := \nf{\Val \{ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t \}}
}
\)
%\rn{Assign}
}
\]
%
The rule chooses an unassigned variable in $\Val$
and assigns it a value that is not $\alpha$-equivalent to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $\FV( \Val )$ and does not add any variables to $\FV( \Val )$,
it can only be applied a finite number of times.
After assigning this value, we normalize all terms in the range of $\Val$.

@@

%%% TYPESETTING: Evil hack!!
%We first state the following properties of $\Val^\AST$,
In the following lemma about $\Val^\AST$,
$\vphantom{\Bigl(_j}\muvar( t ) = \smash{\begin{cases}
  x & \text{if $t$ is of the form $\MU {x}.\; u$} \\[-\jot]
  t & \text{otherwise.}
\end{cases}}$

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\Val^\AST$ is a completion of $\Val$ for a normal model $\M$, the following properties hold:
\begin{enumerate}
\item[\rm (1)] $\Val^\AST \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item[\rm (2)]
$\Val^\AST \ec{x} = \interp{\Val^\AST \ec{y}}{ t }$ for
all subterms $t$ of $\Val^\AST \ec{y}$ with $\muvar(t) = {\Varec{x}}$. 
\item[\rm (3)]
$\Val^\AST \ec{x} \vsim \Val^\AST \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{report}
\begin{proof}
To show 1, we first show that $\Val^\AST$ contains no free variables.
Assume by contradiction that $\Val^\AST$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Single} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\AST)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\AST)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_{\Varec{y}}(\Val^\AST)$ to ${\Varec{y}}$.
By construction, $\Val^\AST \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\AST$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\Val^\AST \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\AST \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

@@

Thus, by induction on the number of applications of the above rule used to obtain $\Val^\AST$,
we have that $\Val^\AST$ satisfies (2) and (3).
\qed
\end{proof}
\end{report}

\begin{theorem}[Solution Soundness]%
\label{thm:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is $\thD$-satisfiable.
\end{theorem}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
%We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where
%all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
%The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
%The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding equalities to $\Fc$.
We consider a normal model $\M$
that interprets wrongly applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\AST$ of the mapping $\Val$ from phase~2.
For the latter, let $\M( x )$ be the value in $\M$ that is $\alpha$-equivalent with $\Val^\AST \ec{x}$ for each variable $x \in \mathcal{T}(\Fc)$.

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.
We show,
by structural induction on $t$,
that $\M( t ) \aequiv \Val^\AST \ec{t}$ for all terms $t \in \mathcal{T}( \Fc )$,
which since $\M$ is normal, implies $\M \models t_1 \teq t_2$.

If $t$ is a variable, then $\M( t ) \vsim \Val^\AST \ec{ t }$ by construction.

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \Val^\AST \ec{ u_1 }, \ldots, \Val^\AST \ec{ u_n } )}$, call this term $t_u$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\Val^\AST$, we have $\Val^\AST \ec{t}$ is a term
%of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
%where $\interp{\Val^\AST \ec{t}}{w_i} = \Val^\AST \ec{u_i}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by our construction of $\Val^\AST$ we have $\Val^\AST \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$ 
where for each $i \in [n]$, 
$\muvar(w_i) = {\Varec{u_i}}$ and thus by Lemma~\ref{lem:model-completion}\PART{2}, 
$\interp{\Val^\AST \ec{t}}{w_i} = \Val^\AST \ec{u_i}$.
For each $i \in [n]$, let $u_\iPrime$ be the $i$\vvthinspace th argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $\interp{t_u}{u_\iPrime} \vsim \Val^\AST \ec{u_i}$.
Thus $\interp{t_u}{u_\iPrime} \vsim \interp{\Val^\AST \ec{t}}{ w_i }$. % for each $i \in [n]$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\M( t ) \vsim t_u \vsim \Val^\AST \ec{t}$, and
we have $\M( t ) \vsim \Val^\AST \ec{t}$.

If $t$ is a selector term of the form $\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\const C_{\negvthinspace\jPrime}\bigl( \const s^1_{\negvthinspace\jPrime}( u ), \ldots, \const s^{n}_{\negvthinspace\jPrime}( u ) \bigr)$ for some $\jPrime$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\Val^\AST$, we have $\Val^\AST \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\interp{\Val^\AST \ec{u} }{ w_i } = \Val^\AST \ec{ \const s^i_{\negvthinspace\jPrime}( u )}$ for $i \in [n]$. 
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by construction $\Val^\AST$ must be of the form
$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
where for each $i \in [n]$,
$\muvar(w_i) = \smash{\Varec{\vphantom{I}\smash{\const s^i_{\negvthinspace\jPrime}( u )}}}$ and thus by Lemma~\ref{lem:model-completion}\PART{2}, 
$\interp{\Val^\AST \ec{u}}{ w_i } = \Val^\AST \ec{ \smash{\const s^i_{\negvthinspace\jPrime}( u )}}$. 
If $j = \jPrime$, then $\M( t )$ is $\alpha$-equivalent with $\interp{\Val^\AST \ec{u}}{w_k}$, which is equal to $\Val^\AST \ec{ \smash{ \const s^k_{\negvthinspace j}( u )}}$,
which is $\Val^\AST \ec{t}$.
If $j \neq \jPrime$, we are free to interpret $\smash{\const s^k_{\negvthinspace j}( \M( u ) )}$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\Val^\AST \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}\PART{3},
pairs of wrongly applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

We now show that all disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
Since $\M( t_1 ) \vsim \Val^\AST \ec{t_1}$ and $\M( t_2 ) \vsim \Val^\AST \ec{t_2}$,
by Lemma~\ref{lem:model-completion}\PART{3}, $\M( t_1 ) \neq \M( t_2 )$, and thus $\M \models t_1 \tneq t_2$.

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcal{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcal{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\AST(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\kern.45em \parallel \kern.45em}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
%\newcommand{\interp}[2]{#1\llbracket #2 \rrbracket}
%%% @ANDY: I prefer a notation that puts more emphasis on #2 (which is, in essence,
%%% the object that is returned) and less on #1
%\newcommand{\interp}[2]{\llbracket #2 \rrbracket_{#1}}
%%% And one that doesn't look so semantic
%\newcommand{\interp}[2]{\lceil #2 \rceil_{#1}}
\newcommand{\interp}[2]{\left< \smash{#2} \right>_{#1}}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\ValC}{\mathcal{A}^\star}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
%\newcommand{\nf}[1]{{{#1}{\downarrow}}}
\newcommand{\nf}[1]{\lfloor#1\rfloor}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

The $\mu$-term $u$
is a \emph{self-similar subterm} of~$t$ if
$u$ is a strict subterm of $t$,
$t$ is of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$,
$u$ it is of the form $\MU y.\; \const{C}( u_1, \ldots, u_n )$,
and $\interp{t}{ t_k } \aequiv \interp{t}{ u_k }$ for all $k \in [n]$.
The term $t$ is \emph{normal} if it does not contain a self-similar subterm
and all of its strict subterms are also normal.
For example, $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal,
because $\MU y.\; \const{C}( y )$ is a self-similar subterm of $t$.
In particular, note that their arguments have the same expansion with respect to $t$,
that is,
$\interp{t}{\MU y.\; \const{C}( y )} =$
$\MU y.\; \const{C}( \interp{t}{y}^{\{y\}} ) =$
$\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$\interp{t}{y} =$
$\MU y.\; \const{C}( \interp{t}{y}^{\{y\}} ) =$
$\MU y.\; \const{C}( y )$.
The $\mu$-term $t$ of the form $\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a self-similar subterm of $t$,
noting that
$\interp{t}{\MU y.\; \const{C}( x )} =
\MU y.\; \const{C}( \interp{t}{x}^{\{y\}} ) =
\MU y.\; \const{C}( \interp{t}{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\{y\}} ) =
\MU y.\; \const{C}( \MU x.\; \const{C}( \interp{t}{\MU y.\; \const{C}( x )}^{\{x,y\}} ) ) =
\MU y.\allowbreak\; \const{C}( \MU x.\; \const{C}( y ) )$
is $\alpha$-equivalent to $\interp{t}{x} = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.
For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$, we can
recursively construct its \emph{normal form} $\nf{t}$,
by replacing all of the self-similar subterms of $t$ with $x$
%and by replacing all of the abnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For example, $\nf{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )} = \MU x.\; \const{C}( x )$.

@@

In the following, we consider only normal models.
When constructing a model $\M$ for $\Ec$,
it remains only to define how $\M$ interprets wrongly applied selector terms and variables.
For the latter, this will be based on the mapping $\Val$ constructed in phase~2 of the calculus.
We compute the \emph{completion} $\ValC$ of $\Val$ for normal model $\M$
by assigning values from $\M$ to unassigned variables in the domain of $\Val$.

@@

We construct $\ValC$ by exhaustively applying the following rule to $\nf{\Val}$:
\[
\hbox{\(
\inferrule{
  {\Varec{x}}\typ{\tau} \in \FV( \Val )
  \quad
  \MU {\Varec{x}}.\; t \aequiv v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_{\Varec{x}}(\Val)
}{
  \Val := \nf{\Val \{ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t \}}
}
\)
%\rn{Assign}
}
\]
%
The rule chooses an unassigned variable in $\Val$
and assigns it a value that is not $\alpha$-equivalent to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $\FV( \Val )$ and does not add any variables to $\FV( \Val )$,
it can only be applied a finite number of times.
After assigning this value, we normalize all terms in the range of $\Val$.

@@

%%% TYPESETTING: Evil hack!!
%We first state the following properties of $\ValC$,
In the following lemma about $\ValC$,
$\vphantom{\Bigl(_j}\muvar( t ) = \smash{\begin{cases}
  x & \text{if $t$ is of the form $\MU {x}.\; u$} \\[-\jot]
  t & \text{otherwise.}
\end{cases}}$

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\ValC$ is a completion of $\Val$ for a normal model $\M$, the following properties hold:
\begin{enumerate}
\item[\rm (1)] $\ValC \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item[\rm (2)]
$\ValC \ec{x} = \interp{\ValC \ec{y}}{ t }$ for
all subterms $t$ of $\ValC \ec{y}$ with $\muvar(t) = {\Varec{x}}$. 
\item[\rm (3)]
$\ValC \ec{x} \vsim \ValC \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{report}
\begin{proof}
To show 1, we first show that $\ValC$ contains no free variables.
Assume by contradiction that $\ValC$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Single} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_{\Varec{y}}(\ValC)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_{\Varec{y}}(\ValC)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_{\Varec{y}}(\ValC)$ to ${\Varec{y}}$.
By construction, $\ValC \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\ValC$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\ValC \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\ValC \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

@@

Thus, by induction on the number of applications of the above rule used to obtain $\ValC$,
we have that $\ValC$ satisfies (2) and (3).
\qed
\end{proof}
\end{report}

\begin{theorem}[Solution Soundness]%
\label{thm:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is $\thD$-satisfiable.
\end{theorem}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
%We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where
%all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
%The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
%The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding equalities to $\Fc$.
We consider a normal model $\M$
that interprets wrongly applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\ValC$ of the mapping $\Val$ from phase~2.
For the latter, let $\M( x )$ be the value in $\M$ that is $\alpha$-equivalent with $\ValC \ec{x}$ for each variable $x \in \mathcal{T}(\Fc)$.

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.
We show,
by structural induction on $t$,
that $\M( t ) \aequiv \ValC \ec{t}$ for all terms $t \in \mathcal{T}( \Fc )$,
which since $\M$ is normal, implies $\M \models t_1 \teq t_2$.

If $t$ is a variable, then $\M( t ) \vsim \ValC \ec{ t }$ by construction.

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \ValC \ec{ u_1 }, \ldots, \ValC \ec{ u_n } )}$, call this term $t_u$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{t}$ is a term
%of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
%where $\interp{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by our construction of $\ValC$ we have $\ValC \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$ 
where for each $i \in [n]$, 
$\muvar(w_i) = {\Varec{u_i}}$ and thus by Lemma~\ref{lem:model-completion}\PART{2}, 
$\interp{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$.
For each $i \in [n]$, let $u_\iPrime$ be the $i$\vvthinspace th argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $\interp{t_u}{u_\iPrime} \vsim \ValC \ec{u_i}$.
Thus $\interp{t_u}{u_\iPrime} \vsim \interp{\ValC \ec{t}}{ w_i }$. % for each $i \in [n]$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\M( t ) \vsim t_u \vsim \ValC \ec{t}$, and
we have $\M( t ) \vsim \ValC \ec{t}$.

If $t$ is a selector term of the form $\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\const C_{\negvthinspace\jPrime}\bigl( \const s^1_{\negvthinspace\jPrime}( u ), \ldots, \const s^{n}_{\negvthinspace\jPrime}( u ) \bigr)$ for some $\jPrime$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\interp{\ValC \ec{u} }{ w_i } = \ValC \ec{ \const s^i_{\negvthinspace\jPrime}( u )}$ for $i \in [n]$. 
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by construction $\ValC$ must be of the form
$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
where for each $i \in [n]$,
$\muvar(w_i) = \smash{\Varec{\vphantom{I}\smash{\const s^i_{\negvthinspace\jPrime}( u )}}}$ and thus by Lemma~\ref{lem:model-completion}\PART{2}, 
$\interp{\ValC \ec{u}}{ w_i } = \ValC \ec{ \smash{\const s^i_{\negvthinspace\jPrime}( u )}}$. 
If $j = \jPrime$, then $\M( t )$ is $\alpha$-equivalent with $\interp{\ValC \ec{u}}{w_k}$, which is equal to $\ValC \ec{ \smash{ \const s^k_{\negvthinspace j}( u )}}$,
which is $\ValC \ec{t}$.
If $j \neq \jPrime$, we are free to interpret $\smash{\const s^k_{\negvthinspace j}( \M( u ) )}$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\ValC \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}\PART{3},
pairs of wrongly applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

We now show that all disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
Since $\M( t_1 ) \vsim \ValC \ec{t_1}$ and $\M( t_2 ) \vsim \ValC \ec{t_2}$,
by Lemma~\ref{lem:model-completion}\PART{3}, $\M( t_1 ) \neq \M( t_2 )$, and thus $\M \models t_1 \tneq t_2$.
